---
title: "Generating_Expression_Correlations"
output: html_document
date: "2023-06-29"
---

The aim of this analysis is to generate correlation plots of the data obtaiend from the expression analysis

Software Versions 

R version 4.2.3 (2023-03-15)
ggplot2 version 3.4.2
dplyr version 1.1.2 
tibble version 3.2.1 
Hmisc version 5.1-0 
corrplot version 0.92
devEMF version 4.4 


Firstly, Remove all files from R's environment

```{r}

rm(list = ls())
```


load the libraries I require


```{r}


library(ggplot2)
library(dplyr)
library(Hmisc)
library(corrplot) 
```



Read in the expression files that were obtained from the dream analysis.  Create a character vector of  all the files in the directory containing the phrase "_Dream_24_06_2023.txt" in their name and save this as "Expression_Filenames"


```{r}

Expression_Filenames <- list.files("/Users/s1790958/Library/CloudStorage/OneDrive-UniversityofEdinburgh/PhD_ Project_Obbard_Lab/DESEQ_Paper-oumie-laptop/UPDATED_DESEQ_FOR_PAPER/Expression_Paper_June_2023", pattern="_Dream_24_06_2023.txt")


Expression_Filenames <- Expression_Filenames [-c(grep("Intercept_|Sex_Effect_Only|Uninfected",Expression_Filenames))]



Expression_File_List <- lapply(Expression_Filenames, function(x) read.table(x, header = TRUE,sep = "\t", row.names = 1) )




names (Expression_File_List) <- Expression_Filenames 
```

Make a vector of the genes that are common across the expression files in our analysis and name this Common_Genes


```{r}

Common_Genes <-  Reduce(intersect, lapply(Expression_File_List, row.names))

```



Now only keep the geens that are present in Common genes from each of the dataframes in "Expression_File_List"


```{r}

#Create an empty list called "Subset_Expression_File_List" to put the expression changes of the genes that will be used in the correlation plots i.e. Common genes

Subset_Expression_File_List <- list()


#Create a for loops to loop through each of the files in "Expression_File_List"


for (i in 1:length( Expression_File_List)) {

  
#Create a dataframe of each file in "All_Expression_All_File"
  
  
Expression_File_List[[i]] <- as.data.frame(Expression_File_List[[i]]) 


#Order each dataframe in   "All_Expression_All_File" by the rownames i.e. the gene names


Expression_File_List[[i]] <- Expression_File_List[[i]] [ order (row.names(Expression_File_List[[i]])), ] 


#Make a subset of the expression changes of the genes found in "Common_Genes" in each dataframe in Expression_File_Listand save this as "Subset_File"


Subset_File <-  subset( Expression_File_List[[i]], row.names(Expression_File_List[[i]] ) %in% Common_Genes)  


#Append each Subset_File to a list called "Subset_Expression_File_List"

Subset_Expression_File_List<- c(Subset_Expression_File_List, list(Subset_File))                                                                                                             
}


#Make the names in "Subset_Expression_File_List" as the names "Expression_File_List"

names (Subset_Expression_File_List) <- names (Expression_File_List)

```

Now only select the LogFC values (i.e. change in expression)
```{r}

#Create an empty list called "LogFC_List" to store the LogFC values


LogFC_List <- list()



#each file in Subset_Expression_File_List


for (i in 1:length( Subset_Expression_File_List)) {
  

#Get the logFC values in each expression file and save it as "LogFC_df"
  
  
LogFC_df <- Subset_Expression_File_List[[i]]$logFC


#make a dataframe of the LogFC_df


LogFC_df <- as.data.frame(LogFC_df)

#Name the column "LogFC"

 
names (LogFC_df) <-  c("LogFC")


#Put the LogFC_df into a list called "LogFC_List"


LogFC_List  <-c(LogFC_List , list(LogFC_df) )

}


#Give the "LogFC_List" the same names as the "Subset_Expression_File_List"


names (LogFC_List) <- names (Subset_Expression_File_List)


```



```{r}


#Turn the LogFC_List into. dataframe and call it "Expression_File_df"

Expression_File_df<- as.data.frame(LogFC_List)


#Name the everything daatframe as the names of the LogFC_List


names (Expression_File_df) <- names (LogFC_List)

#Now remove "_Effect_Dream_24_06_2023.txt" or "_Interaction_Dream_24_06_2023.txt from the names of the columns in "Expression_File_df"


names (Expression_File_df) <- gsub ("_Effect_Dream_24_06_2023.txt|_Interaction_Dream_24_06_2023.txt", " ", names (Expression_File_df))

#Now remove "_" from the names of the columns in "Expression_File_df"


names (Expression_File_df) <- gsub ("_", " ", names (Expression_File_df))

#Rename each " Sex" to ":Sex"

names (Expression_File_df) <- gsub (" Sex", ":Sex", names (Expression_File_df))




```


Create a dataframe with only the main virus effects on gene expression

```{r}

Main_Virus_Effect <- Expression_File_df[-c(grep ("Sex", names (Expression_File_df)))]
```



Create a dataframe with only the virus:sex effects on gene expression

```{r}

Sex_Virus_Interaction_Effect <- Expression_File_df[(grep (":Sex", names(Expression_File_df)))]


```



Generate correlations "pearson's" correlations using rcorr

```{r}
#Make a correlation matrix of the main virus effects on gene expression

Corr_Matrix_Main_Virus_Effect <- rcorr(as.matrix (Main_Virus_Effect),type=c("pearson"))

#Replace the NA with 0

Corr_Matrix_Main_Virus_Effect$P[is.na(Corr_Matrix_Main_Virus_Effect$P)] <- 0
Corr_Matrix_Main_Virus_Effect$r[is.na(Corr_Matrix_Main_Virus_Effect$r)] <- 0
Corr_Matrix_Main_Virus_Effect$n[is.na(Corr_Matrix_Main_Virus_Effect$n)] <- 0

#Make a correlation matrix of the virus:sex effects on gene expression

Corr_Matrix_Sex_Virus_Interaction_Effect <- rcorr(as.matrix (Sex_Virus_Interaction_Effect),type=c("pearson"))

#Replace the NA with 0

Corr_Matrix_Sex_Virus_Interaction_Effect$P[is.na(Corr_Matrix_Sex_Virus_Interaction_Effect$P)] <- 0
Corr_Matrix_Sex_Virus_Interaction_Effect$r[is.na(Corr_Matrix_Sex_Virus_Interaction_Effect$r)] <- 0
Corr_Matrix_Sex_Virus_Interaction_Effect$n[is.na(Corr_Matrix_Sex_Virus_Interaction_Effect$n)] <- 0

```



Create a plot of the "Corr_Matrix_Main_Virus_Effect" and save this as "Main_Virus_Effect_Correlations_29_06_2023.emf"


```{r}


LabelCol = c("black")
emf(file= "Main_Effect_Correlations_Plot_28_06_2023.emf", height =20, width = 20, emfPlus = FALSE)
 
corrplot(Corr_Matrix_Main_Virus_Effect$r, type="upper", p.mat = Corr_Matrix_Main_Virus_Effect$P,  insig = "blank", method = "number", sig.level = 0.001, addgrid = TRUE, order = "alphabet", number.cex=3, tl.cex =3.5, cl.cex = 3, tl.col=LabelCol, col=colorRampPalette(c("firebrick","grey50", "blue"))(200))
dev.off()
```



Create a plot of the "Corr_Matrix_Main_Virus_Effect" and save this as "Virus_Sex_Interaction_Effect_Correlations_29_06_2023.emf"


```{r}



LabelCol = c("black")
emf(file= "Virus_Sex_Interaction_Effect_Correlations_Plot_28_06_2023.emf", height =10, width = 10, emfPlus = FALSE)
 
corrplot(Corr_Matrix_Sex_Virus_Interaction_Effect$r, type="upper", p.mat = Corr_Matrix_Sex_Virus_Interaction_Effect$P,  insig = "blank", method = "number", sig.level = 0.001, addgrid = TRUE, order = "alphabet", number.cex=2, tl.cex =2, cl.cex = 2, tl.col=LabelCol, col=colorRampPalette(c("firebrick","grey50", "blue"))(200))
dev.off()
```
