---
title: "Generating Volcano Plots from Expression Analyses"
output: html_document
date: "2023-06-29"
---

The aim of this analysis is to generate Volcano plots of the data obtaiend from the expression analysis

Software Versions 

R version 4.2.3 (2023-03-15)
ggplot2 version 3.4.2
ggpubr version 0.6.0
dplyr version 1.1.2 
tibble version 3.2.1 
EnhancedVolcano version 1.16.0 
devEMF version 4.4 


Firstly, Remove all files from R's environment

```{r}

rm(list = ls())
```


Load the required libraries 

```{r}


library(ggplot2)
library(ggpubr)
library(dplyr)
library(tibble) 
library(EnhancedVolcano) 
library(devEMF)


```


Read in the expression files that were obtained from the dream analysis. These files all have the phrase "_Dream_24_06_2023.txt" in their name. Create a character vector of  all the files in the directory containing the phrase "_Dream_24_06_2023.txt" in their name and save this as "Expression_Filenames"


```{r}


Expression_Filenames <- list.files("/Users/s1790958/Library/CloudStorage/OneDrive-UniversityofEdinburgh/PhD_ Project_Obbard_Lab/DESEQ_Paper-oumie-laptop/UPDATED_DESEQ_FOR_PAPER/Expression_Paper_June_2023", pattern="_Dream_24_06_2023.txt")



# Remove the file names with 'Intercept" or "Only_" in their name as we are not interested in those


Expression_Filenames <- Expression_Filenames [-c(grep("Intercept_|Sex_Effect_Only",Expression_Filenames))]


#Read all the files by using the lapply (list apply) command. Also, remove the phrase "_Dream_24_06_2023.txt"  to make them shorter.

Expression_File_List <- lapply(Expression_Filenames, function(x) read.table(x, header = TRUE,sep = "\t", row.names = 1) )


names(Expression_File_List) <- gsub("_Dream_24_06_2023.txt","", Expression_Filenames)


```


Now put the all the contents of the Expression_File_List in the global environment


```{r}

list2env (Expression_File_List, envir = .GlobalEnv)

```


We will draw a plot for of the main effects. i.e  for the effects of Sex, galbut virus, Motts Mill virus, nora virus, DCV, DAV, Thika virus, Brandeis virus , Craigies Hill, Bloomfield virus, and DMelSV.


```{r}

#Sex Effect Uninfected Plot 

Sex_Effect_Uninfected_plot <- EnhancedVolcano(Sex_Effect_Uninfected,
    lab = rownames(Sex_Effect_Uninfected),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Sex',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 20),
pointSize = 6.0,
    labSize = 8.0)




#galbut virus Effect Plot 

galbut_virus_Effect_plot <- EnhancedVolcano(galbut_virus_Effect,
    lab = rownames(galbut_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'galbut virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 11),
pointSize = 6.0,
    labSize = 8.0)



#Motts Mill virus Effect Plot 

Motts_Mill_virus_Effect_plot <- EnhancedVolcano(Motts_Mill_virus_Effect,
    lab = rownames(Motts_Mill_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Motts Mill virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 20),
pointSize = 6.0,
    labSize = 8.0)




#nora virus Effect Plot 

nora_virus_Effect_plot <- EnhancedVolcano(nora_virus_Effect,
    lab = rownames(nora_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'nora virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 20),
pointSize = 6.0,
    labSize = 8.0)




#DCV Effect Uninfected Plot 

DCV_Effect_plot <- EnhancedVolcano(DCV_Effect,
    lab = rownames(DCV_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'DCV',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
  xlim=c(-2, 2),
ylim=c(0, 4.5),
pointSize = 6.0,
    labSize = 8.0)



#DAV Effect Uninfected Plot 

DAV_Effect_plot <- EnhancedVolcano(DAV_Effect,
    lab = rownames(DAV_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'DAV',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 7.5),
pointSize = 6.0,
    labSize = 8.0)



#Thika virus Effect Uninfected Plot 

Thika_virus_Effect_plot <- EnhancedVolcano(Thika_virus_Effect,
    lab = rownames(Thika_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Thika virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 9),
pointSize = 4.0,
    labSize = 6.0)



#Craigies Hill virus Effect Uninfected Plot 

Craigies_Hill_virus_Effect_plot <- EnhancedVolcano(Craigies_Hill_virus_Effect,
    lab = rownames(Craigies_Hill_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Craigies Hill virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 9),
pointSize = 6.0,
    labSize = 8.0)



#Brandeis virus Effect Uninfected Plot 

Brandeis_virus_Effect_plot <- EnhancedVolcano(Brandeis_virus_Effect,
    lab = rownames(Brandeis_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Brandeis virus',
        titleLabSize = 40,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 45,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 20),
pointSize = 6.0,
    labSize = 8.0)



#Bloomfield virus Effect Uninfected Plot 

Bloomfield_virus_Effect_plot <- EnhancedVolcano(Bloomfield_virus_Effect,
    lab = rownames(Bloomfield_virus_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Bloomfield virus',
        titleLabSize =25,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 30,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 7),
pointSize = 4.0,
    labSize = 6.0)




#DMelSV Effect Uninfected Plot 

DMelSV_Effect_plot <- EnhancedVolcano(DMelSV_Effect,
    lab = rownames(DMelSV_Effect),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'DMelSV',
        titleLabSize = 25,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 30,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
ylim=c(0, 3),
pointSize = 4.0,
    labSize = 6.0)



```


Combine the main effects plots i.e. for Sex effect,  galbut virus, Motts Mill virus, nora virus, DCV, DAV, Thika virus,   Brandeis virus , and Craigies Hill virus using ggarrage and save this save this as "figure". Then export "figure" it into an emf file called "Main_Effects_Volcano_Plot_28_06_2023.emf".

```{r}


emf(file="Main_Effects_Volcano_Plot_28_06_2023.emf", width=21, height=25, emfPlus = FALSE)

figure <- ggarrange(Sex_Effect_Uninfected_plot, galbut_virus_Effect_plot, Motts_Mill_virus_Effect_plot, nora_virus_Effect_plot, DCV_Effect_plot, DAV_Effect_plot, Thika_virus_Effect_plot ,  Brandeis_virus_Effect_plot , Craigies_Hill_virus_Effect_plot ,
                    labels = c("A", "B", "C", "D", "E", "F", "G", "H", "I"), font.label = list(size = 40, color = "black"),
                    
                    ncol = 3, nrow = 3)
figure

dev.off()
```


Combine the main effects plots of Bloomfield virus and DMelSV using ggarrage and save this save this as "figure", and export into in an emf file called "Bloomfield_virus_DMelSV_Volcano_Plot_28_06_2023.emf"

```{r}

emf(file="Bloomfield_virus_DMelSV_Volcano_Plot_28_06_2023.emf", width=10, height=6, emfPlus = FALSE)

figure <- ggarrange( Bloomfield_virus_Effect_plot, DMelSV_Effect_plot,
                    labels = c("A", "B"), font.label = list(size = 30, color = "black"),
                    
                    ncol = 2, nrow = 1)
figure

dev.off()
```


We will draw a plot for virus:sex interactions for gakbut virus, Motts Mill virus, nora virus, DCV, DAV, and Thika virus.

```{r}

#Draw Plots for each of the virus:Sex Effects

#galbut virus:Sex Plot 

galbut_virus_Sex_Interaction_plot <- EnhancedVolcano(galbut_virus_Sex_Interaction,
    lab = rownames(galbut_virus_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'galbut virus:Sex',
        titleLabSize = 30,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 20),
pointSize = 4.0,
    labSize = 6.0)



#Motts Mill virus:Sex Plot 

Motts_Mill_virus_Sex_Interaction_plot <- EnhancedVolcano(Motts_Mill_virus_Sex_Interaction,
    lab = rownames(Motts_Mill_virus_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Motts Mill virus:Sex',
    titleLabSize = 30,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
     xlim=c(-4, 4),
#ylim=c(0, 500),
pointSize = 4.0,
    labSize = 6.0)


#nora virus:Sex Plot 

nora_virus_Sex_Interaction_plot <- EnhancedVolcano(nora_virus_Sex_Interaction,
    lab = rownames(nora_virus_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'nora virus:Sex',
        titleLabSize = 30,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
    # xlim=c(-4, 4),
#ylim=c(0, 500),
pointSize = 4.0,
    labSize = 6.0)



#DCV:Sex Plot 

DCV_Sex_Interaction_plot <- EnhancedVolcano(DCV_Sex_Interaction,
    lab = rownames(DCV_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'DCV:Sex',
        titleLabSize = 30,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
  xlim=c(-3.5, 3),
ylim=c(0, 3),
pointSize = 4.0,
    labSize = 6.0)




#DAV:Sex Plot 

DAV_Sex_Interaction_plot <- EnhancedVolcano(DAV_Sex_Interaction,
    lab = rownames(DAV_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'DAV:Sex',
        titleLabSize = 30,
   subtitle = NULL,
  caption = NULL, 
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
   xlim=c(-3, 3),
ylim=c(0, 11),
pointSize = 4.0,
    labSize = 6.0)



#Thika virus:Sex Plot 

Thika_virus_Sex_Interaction_plot <- EnhancedVolcano(Thika_virus_Sex_Interaction,
    lab = rownames(Thika_virus_Sex_Interaction),
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Thika virus:Sex',
        titleLabSize = 30,
   subtitle = NULL,
  caption = NULL,
    pCutoff = 0.001,
     FCcutoff = 0.5,
    axisLabSize = 35,
  ylab = expression('-Log'[10]*italic("P")),
    legendPosition = 'none',
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 2, 
   cutoffLineType = 'blank',
   #  xlim=c(-3.5, 3.5),
ylim=c(0, 9),
pointSize = 4.0,
    labSize = 6.0)

```


Combine the virus:Sex plots and export it in an emf file called "Virus_Sex_Interaction_Volcano_Plot_28_06_2023.emf"

```{r}

emf(file="Virus_Sex_Interaction_Volcano_Plot_28_06_2023.emf", width=16, height=16, emfPlus = FALSE)

figure <- ggarrange(galbut_virus_Sex_Interaction_plot, Motts_Mill_virus_Sex_Interaction_plot, nora_virus_Sex_Interaction_plot, DCV_Sex_Interaction_plot, DAV_Sex_Interaction_plot, Thika_virus_Sex_Interaction_plot,
                    labels = c("A", "B", "C", "D", "E", "F"), font.label = list(size = 30, color = "black"),
                    
                    ncol = 3, nrow = 2)
figure

dev.off()
```


